\section{Canonical Failure Modes}

This section presents three canonical failure modes in which predictive models perform
well under observation yet fail under intervention. Each case follows the same pattern:
a model interface induces equivalence classes over trajectories that are harmless during
training but unstable under admissible interventions. In all cases, the resulting
failure is predictable prior to deployment using the diagnostic described in
Section~6.

\subsection{Sim-to-Real Robotics}

In sim-to-real robotics, a simulator is often validated by its ability to reproduce
observed rollouts under a fixed policy. At the simulator interface, many distinct
physical trajectories may be collapsed into a single abstract state, such as a
low-dimensional pose or contact representation. Under observation, this reduction can
be predictively adequate.

Deployment introduces actuation. Control inputs act on the physical system, not on the
abstract state. Small differences in contact dynamics, friction, compliance, or timing
can separate trajectories that were equivalent at the simulator interface. As a result,
the controller encounters outcomes that the simulator could not distinguish.

\paragraph{Diagnosis.}
The simulator interface induces an equivalence class over trajectories that is stable
under observation but unstable under actuation. The reduction is therefore an unfaithful
cut at the physics abstraction layer.

\subsection{Offline Reinforcement Learning}

Offline reinforcement learning trains policies from logged trajectories generated under
one or more behavior policies. Evaluation typically assumes that the learned policy will
operate within the support of the logged data. At the model interface, trajectories are
grouped by a state representation learned or assumed during training.

A change in policy constitutes an intervention. Even a small policy change can induce
actions that separate trajectories previously treated as equivalent at the interface.
Because the offline model cannot distinguish these trajectories, the induced behavior
can fall outside the regime in which predictions were valid.

\paragraph{Diagnosis.}
The state representation collapses histories that are observationally equivalent under
the behavior policy but distinguishable under the learned policy. The equivalence is not
stable under intervention, yielding an unfaithful cut at the state definition.

\subsection{Causal Inference in Medicine}

In medical decision-making, predictive models are often trained on observational patient
data to estimate outcomes. At the model interface, patients with different underlying
histories may be mapped to the same representation based on observed variables. This can
yield accurate predictions of outcomes under observed treatment distributions.

Treatment selection is an intervention. A chosen treatment can separate patient
trajectories that were indistinguishable at the interface. Latent factors, unobserved
progression, or prior interventions can cause identical interface states to respond
differently to the same action.

\paragraph{Diagnosis.}
The patient representation conflates histories that are equivalent under observation but
distinct under intervention. The resulting model is observationally adequate but not
interventionally valid, indicating an unfaithful cut.